---
title: "Week2 - garden of forking data"
author: "P. Melloy"
date: "13/08/2020"
output: html_document
---

```{r install&load_Libraries}
library(rethinking)
```

Baysian inference is the  
$ProbabilityOfTheData * \frac{PriorProbability}{NormalisationToAProbability}$

## Code from the book
```{r grid_aprox}
p_grid <- seq(from = 0, to = 1, length.out = 100)
prob_p <- rep(1,100)
prob_data <- dbinom(6, size = 9, prob = p_grid)
posterior <- prob_data * prob_p
posterior <- posterior/sum(posterior)

plot(p_grid)
plot(prob_p) # plot the prior
plot(prob_data)
points(posterior, col = "blue")
```

> Confidence intervals or any intervals have no meaning and there could be infinite number of intervals.  

```{r chunk2.3}
n1 <- 1000
# define grid
p_grid <- seq( from=0 , to=1 , length.out= n1 )
# define prior
#prior <- rep( 1 , n1 )
#prior <- ifelse( p_grid < 0.5 , 0 , 1 )
prior <- exp( -5*abs( p_grid - 0.5 ) )


# compute likelihood at each value in grid
likelihood <- dbinom( 6 , size=9 , prob=p_grid )
# compute product of likelihood and prior
unstd.posterior <- likelihood * prior
# standardize the posterior, so it sums to 1
posterior <- unstd.posterior / sum(unstd.posterior)

plot( p_grid , posterior , type="b" ,
xlab="probability of water" , ylab="posterior probability" )
mtext(paste(n1, "points" ))

```


## Installing 'rethinking' package
Note: I have already followed the steps to install Rstan.
If you have not installed stan for R the visit the following link and follow the instructions.
[https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started](https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started)
I am using a Windows 10 machine

```{r install_rethinking, eval=FALSE, echo=TRUE}
#This chunk is not evaluated
install.packages(c("coda","mvtnorm","devtools"))
library(devtools)
devtools::install_github("rmcelreath/rethinking")
```

```{r chunk_2.6}
globe.qa <- quap(
alist(
W ~ dbinom( W+L ,p) , # binomial likelihood
p ~ dunif(0,1) # uniform prior
) ,
data=list(W=6,L=3) )
# display summary of quadratic approximation
precis( globe.qa )

```

```{r chunk2.7}
# analytical calculation 2.7
W <- 6
L <- 3
curve( dbeta( x , W+1 , L+1 ) , from=0 , to=1 )
# quadratic approximation
curve( dnorm( x , 0.67 , 0.16 ) , lty=2 , add=TRUE )
```

```{r chunk2-8}
n_samples <- 1000
p <- rep( NA , n_samples )
p[1] <- 0.5
W <- 6
L <- 3
for ( i in 2:n_samples ) {
p_new <- rnorm( 1 , p[i-1] , 0.1 )
if ( p_new < 0 ) p_new <- abs( p_new )
if ( p_new > 1 ) p_new <- 2 - p_new
q0 <- dbinom( W , W+L , p[i-1] )
q1 <- dbinom( W , W+L , p_new )
p[i] <- ifelse( runif(1) < q1/q0 , p_new , p[i-1] )
}
dens(p, xlim=c(0,1))
curve( dbeta( x , W+1 , L+1 ) , lty=2 , add=TRUE )
```

So line 104 says if the new mean `q1` is a larger number than the prior mean then use the p_new.
If not use the old prior mean 0.5



## Exercises  

#### Easy questions
2E1. Which of the expressions below correspond to the statement: the probability of rain on Monday?  
   (1) Pr(rain)  
   (2) **Pr(rain|Monday)**  
(3) Pr(Monday|rain)  
(4) **Pr(rain, Monday)/ Pr(Monday)**  

<br> <!--HTML line break-->

2E2. Which of the following statements corresponds to the expression: Pr(Monday|rain)?  
(1) The probability of rain on Monday.  
(2) The probability of rain, given that it is Monday.  
(3) **The probability that it is Monday, given that it is raining.**  
(4) The probability that it is Monday and that it is raining.  

<br> <!--HTML line break-->

2E3. Which of the expressions below correspond to the statement: the probability that it is Monday,
given that it is raining?  
**(1) Pr(Monday|rain)**  
(2) Pr(rain|Monday)  
(3) Pr(rain|Monday) Pr(Monday)  
**(4) Pr(rain|Monday) Pr(Monday)/ Pr(rain)**  
(5) Pr(Monday|rain) Pr(rain)/ Pr(Monday)  

<br> <!--HTML line break-->

2E4. The Bayesian statistician Bruno de Finetti (1906–1985) began his book on probability theory
with the declaration: “PROBABILITY DOES NOT EXIST.” The capitals appeared in the original, so
I imagine de Finetti wanted us to shout this statement. What he meant is that probability is a device
for describing uncertainty from the perspective of an observer with limited knowledge; it has no
objective reality. Discuss the globe tossing example from the chapter, in light of this statement. What
does it mean to say “the probability of water is 0.7”?  

*This really means it is likely that the percentage of water covering the globe surface is 70 %, (or the proportion as 0.7)*.
*Although by saying probability it infers some uncertainty about it, however this is given as the most likley estimate, uncertainty is not given through the variance or standard deviation of the measurement.*
*So by saying that the probability is 0.7 we are saying on average 7 throws out of 10 will land on water.*

<br> <!--HTML line break-->

#### Medium questions  
2M1. Recall the globe tossing model from the chapter. Compute and plot the grid approximate
posterior distribution for each of the following sets of observations. In each case, assume a uniform
prior for p.  
(1) W, W, W  

```{r Q_2M1.1}
n1 <- 1000
# define grid
p_grid <- seq( from=0 , to=1 , length.out= n1 )

# define uniform prior
prior <- rep( 1 , n1 )

# compute likelihood at each value in grid
likelihood <- dbinom( 3 , size=3 , prob=p_grid )
# compute product of likelihood and prior
unstd.posterior <- likelihood * prior
# standardize the posterior, so it sums to 1
posterior <- unstd.posterior / sum(unstd.posterior)

plot( p_grid , posterior , type="b" ,
xlab="probability of water" , ylab="posterior probability" )
mtext(paste(n1, "points" ))
```

<br> <!--HTML line break-->

(2) W, W, W, L  
```{r Q_2M1.2}
n1 <- 1000
# define grid
p_grid <- seq( from=0 , to=1 , length.out= n1 )

# define uniform prior
prior <- rep( 1 , n1 )

# compute likelihood at each value in grid
likelihood <- dbinom( 3 , size=4 , prob=p_grid )
# compute product of likelihood and prior
unstd.posterior <- likelihood * prior
# standardize the posterior, so it sums to 1
posterior <- unstd.posterior / sum(unstd.posterior)

plot( p_grid , posterior , type="b" ,
xlab="probability of water" , ylab="posterior probability" )
mtext(paste(n1, "points" ))
```

<br> <!--HTML line break-->

(3) L, W, W, L, W, W, W  
```{r Q_2M1.3}
n1 <- 1000
# define grid
p_grid <- seq( from=0 , to=1 , length.out= n1 )

# define uniform prior
prior <- rep( 1 , n1 )

# compute likelihood at each value in grid
likelihood <- dbinom( 5 , size=7 , prob=p_grid )
# compute product of likelihood and prior
unstd.posterior <- likelihood * prior
# standardize the posterior, so it sums to 1
posterior <- unstd.posterior / sum(unstd.posterior)

plot( p_grid , posterior , type="b" ,
xlab="probability of water" , ylab="posterior probability" )
mtext(paste(n1, "points" ))
```

<br> <!--HTML line break-->

2M2. Now assume a prior for p that is equal to zero when p < 0.5 and is a positive constant when
p ≥ 0.5. Again compute and plot the grid approximate posterior distribution for each of the sets of
observations in the problem just above.  

(1) W, W, W  

```{r Q_2M2.1}
n1 <- 1000
# define grid
p_grid <- seq( from=0 , to=1 , length.out= n1 )

# define uniform prior for probabilities over 0.5, and zero below 0.5
prior <- ifelse( p_grid < 0.5 , 0 , 1 )


# compute likelihood at each value in grid
likelihood <- dbinom( 3 , size=3 , prob=p_grid )
# compute product of likelihood and prior
unstd.posterior <- likelihood * prior
# standardize the posterior, so it sums to 1
posterior <- unstd.posterior / sum(unstd.posterior)

plot( p_grid , posterior , type="b" ,
xlab="probability of water" , ylab="posterior probability" )
mtext(paste(n1, "points" ))
```

<br> <!--HTML line break-->


(2) W, W, W, L  
```{r Q_2M2.2}
n1 <- 1000
# define grid
p_grid <- seq( from=0 , to=1 , length.out= n1 )

# define uniform prior for probabilities over 0.5, and zero below 0.5
prior <- ifelse( p_grid < 0.5 , 0 , 1 )

# compute likelihood at each value in grid
likelihood <- dbinom( 3 , size=4 , prob=p_grid )
# compute product of likelihood and prior
unstd.posterior <- likelihood * prior
# standardize the posterior, so it sums to 1
posterior <- unstd.posterior / sum(unstd.posterior)

plot( p_grid , posterior , type="b" ,
xlab="probability of water" , ylab="posterior probability" )
mtext(paste(n1, "points" ))
```

<br> <!--HTML line break-->


(3) L, W, W, L, W, W, W  
```{r Q_2M2.3}
n1 <- 1000
# define grid
p_grid <- seq( from=0 , to=1 , length.out= n1 )

# define uniform prior for probabilities over 0.5, and zero below 0.5
prior <- ifelse( p_grid < 0.5 , 0 , 1 )

# compute likelihood at each value in grid
likelihood <- dbinom( 5 , size=7 , prob=p_grid )
# compute product of likelihood and prior
unstd.posterior <- likelihood * prior
# standardize the posterior, so it sums to 1
posterior <- unstd.posterior / sum(unstd.posterior)

plot( p_grid , posterior , type="b" ,
xlab="probability of water" , ylab="posterior probability" )
mtext(paste(n1, "points" ))
```

<br> <!--HTML line break-->


2M3. Suppose there are two globes, one for Earth and one for Mars. The Earth globe is 70% covered
in water. The Mars globe is 100% land. Further suppose that one of these globes—you don’t know
which—was tossed in the air and produced a “land” observation. Assume that each globe was equally
likely to be tossed. Show that the posterior probability that the globe was the Earth, conditional on
seeing “land” (Pr(Earth|land)), is 0.23.  

$Pr(land|Earth) = 1 - 0.7 = 0.3$  
$Pr(land|Mars) = 1$  
$Pr(Earth) = 0.5$  
$Pr(Mars) = 0.5$  
$Pr(land) = Pr(land|Earth)*Pr(Earth) + Pr(land|Mars)*Pr(Mars)$  
```{r}
paste("Pr(land) =",
      (0.3 * 0.5) + (1 * 0.5))
```

<br> <!--HTML line break-->


$Pr(Earth|land) = \frac{Pr(land|Earth)*Pr(Earth)}{Pr(land)}$  
```{r}
paste("Pr(Earth|land) =",
      (0.3 * 0.5)/0.65)

```

*Disclaimer: I needed help so I googled for help*

<br> <!--HTML line break-->


2M4. Suppose you have a deck with only three cards. Each card has two sides, and each side is either
black or white. One card has two black sides. The second card has one black and one white side. The
third card has two white sides. Now suppose all three cards are placed in a bag and shuffled. Someone
reaches into the bag and pulls out a card and places it flat on a table. A black side is shown facing up,
but you don’t know the color of the side facing down. Show that the probability that the other side is
also black is 2/3. Use the counting method (Section 2 of the chapter) to approach this problem. This
means counting up the ways that each card could produce the observed data (a black side facing up
on the table).  

Probability of  
top/bottom  
W/W = 0  
W/W = 0  
B/W = 1  
W/B = 0  
B/B = 1  
B/B = 1  

*We can see that where the top card is black and the bottom side is black occurs 2 times out of the 6(5) equally likely options.*
*However given we see the top is black we have data to eliminate any options where white would be on top.*
*Therefore 2/3*

<br> <!--HTML line break-->


2M5. Now suppose there are four cards: B/B, B/W, W/W, and another B/B. Again suppose a card is
drawn from the bag and a black side appears face up. Again calculate the probability that the other
side is black.  

Probability of  
top/bottom  
B/B = 1  
B/B = 1  
B/W = 1  
W/B = 0  
W/W = 0  
W/W = 0  
B/B = 1
B/B = 1

*We can see that where the top card is black and the bottom side is black occurs 4 times out of the 8 equally likely options.*
*However given we see the top is black we have data to eliminate any options where white would be on top.*
*Therefore 4/5*


<br> <!--HTML line break-->


2M6. Imagine that black ink is heavy, and so cards with black sides are heavier than cards with white
sides. As a result, it’s less likely that a card with black sides is pulled from the bag. So again assume
there are three cards: B/B, B/W, and W/W. After experimenting a number of times, you conclude that
for every way to pull the B/B card from the bag, there are 2 ways to pull the B/W card and 3 ways to
pull the W/W card. Again suppose that a card is pulled and a black side appears face up. Show that
the probability the other side is black is now 0.5. Use the counting method, as before.

Probability of  
top/bottom  
W/W = 0 * 3  
W/W = 0 * 3  
B/W = 1 * 2  
W/B = 0 * 2  
B/B = 1  
B/B = 1  

*We can see that where the top card is black and the bottom side is black occurs 2 times out of the 12.*
*However given we see the top is black we have data to eliminate any options where white would be on top.*
*Therefore 2/(2 + 2x1) = 1/2*

<br> <!--HTML line break-->


2M7. Assume again the original card problem, with a single card showing a black side face up. Before
looking at the other side, we draw another card from the bag and lay it face up on the table. The face
that is shown on the new card is white. Show that the probability that the first card, the one showing
a black side, has black on its other side is now 0.75. Use the counting method, if you can. Hint: Treat
this like the sequence of globe tosses, counting all the ways to see each observation, for each possible
first card.  

Probability of  
top/bottom  
W/W = 0  
W/W = 0  
B/W = 1  
W/B = 0  
B/B = 1  
B/B = 1  



### Hard  
2H1. Suppose there are two species of panda bear. Both are equally common in the wild and live
in the same places. They look exactly alike and eat the same food, and there is yet no genetic assay
capable of telling them apart. They differ however in their family sizes. Species A gives birth to twins
10% of the time, otherwise birthing a single infant. Species B births twins 20% of the time, otherwise
birthing singleton infants. Assume these numbers are known with certainty, from many years of field
research.
Now suppose you are managing a captive panda breeding program. You have a new female panda
of unknown species, and she has just given birth to twins. What is the probability that her next birth
will also be twins?  

$Pr(PandaSpecies_A) = 0.5$
$Pr(PandaSpecies_B) = 1 - Pr(PandaSpecies_A) = 0.5$
$Pr(twins|PandaSpecies_A) = \frac{Pr(PandaSpecies_A|twins)Pr(twins)}{Pr(PandaSpeciesA)}$
$Pr(twins|PandaSpecies_A) = 0.1$
$Pr(twins|PandaSpecies_B) = 0.2$
Calculate the marginal probability of twins (probability of twins irrespective of any other outcome)
$Pr(twins) = Pr(twins|PandaSpecies_A)*Pr(PandaSpecies_A) + Pr(twins|PandaSpecies_B)*Pr(PandaSpecies_B)$
```{r probabilityOfTwins}
paste("Pr(twins) = appox",
      (0.1 * 0.5)+(0.2 * 0.5))
```

The probability of two sets of twins is = $twins^2$
```{r probabilityOfTwoTwins}
paste("Pr(twins)^2 =",
      ((0.1 * 0.5)+(0.2 * 0.5))^2)

paste("Pr(twins)^2 =",
      (0.166667)^2)


# Dans
prior = c(0.5, 0.5)
likelihood = c(0.1, 0.2)
posterior_unst = likelihood * prior
posterior = posterior_unst / sum(posterior_unst)
# probability next birth is twins:
posterior[1] * 0.1 + posterior[2] * 0.2

```


<br> <!--HTML line break-->

2H2. Recall all the facts from the problem above. Now compute the probability that the panda we
have is from species A, assuming we have observed only the first birth and that it was twins.  

$Pr(PandaSpecies_A|twins)=\frac{Pr(twins|PandaSpecies_A)Pr(PandaSpecies_A)}{Pr(twins)}$
```{r}
paste("Pr(PandaSpecies_A|twins) =",
      (0.1 * 0.5)/(0.15))

paste("Pr(PandaSpecies_A|twins) =",
      (0.1 * 0.5)/(0.166667))


# 2H2 - Dans
prior1 = c(0.5, 0.5) # A, B
likelihood1 = c(0.1, 0.2) #A, B
posterior_unst1 = likelihood1 * prior1
posterior1 = posterior_unst / sum(posterior_unst)
# probability she is species A
posterior1[1]

```


<br> <!--HTML line break-->

2H3. Continuing on from the previous problem, suppose the same panda mother has a second birth
and that it is not twins, but a singleton infant. Compute the posterior probability that this panda is
species A.  

First lets find the $Pr(singleton)$  
$Pr(singleton) = Pr(singleton|PandaSpecies_A)*Pr(PandaSpecies_A) + Pr(singleton|PandaSpecies_B)*Pr(PandaSpecies_B)$
```{r probabilityOfsingle}
# I have not updated my prior here Probability of pandaSpecies_A  = 0.33333
# I have not updated my prior here Probability of pandaSpecies_B  = 0.66666
paste("Pr(single) = appox",
      (0.9 * 0.5)+(0.8 * 0.5))

paste("Pr(single) = appox",
      (0.9 * 0.3333)+(0.8 * 0.66666))
```
Which also equals 1 - $Pr(twins)$

What is the probability of PanderSpecies-A given a singleton
$Pr(PandaSpecies_A|singleton)=\frac{Pr(singleton|PandaSpecies_A)Pr(PandaSpecies_A)}{Pr(singleton)}$  
```{r}
# I have not updated my prior here Probability of pandaSpecies_A  = 0.33333
paste("Pr(PandaSpecies_A|single) =",
      (0.9 * 0.5)/(0.85))

paste("Pr(PandaSpecies_A|single) =",
      (0.9 * 0.33333)/(0.85))
```

I feel like I might be on the wrong track here
Add up the probabilities and divide by the sum of all probabilities in each of the sum of all probabilities....
```{r}
(0.85+0.52944)/2

```
```{r}
# 2H3 Dans
prior2 = posterior1
likelihood2 = 1 - likelihood1
posterior_unst2 = prior2 * likelihood2
posterior2 = posterior_unst2 / sum(posterior_unst2)
posterior2
# Prob of singleton given species A and twins
posterior2[1]

```



<br> <!--HTML line break-->

2H4. A common boast of Bayesian statisticians is that Bayesian inference makes it easy to use all of
the data, even if the data are of different types.
So suppose now that a veterinarian comes along who has a new genetic test that she claims can
identify the species of our mother panda. But the test, like all tests, is imperfect. This is the information
you have about the test:
• The probability it correctly identifies a species A panda is 0.8.
• The probability it correctly identifies a species B panda is 0.65.
The vet administers the test to your panda and tells you that the test is positive for species A. First
ignore your previous information from the births and compute the posterior probability that your
panda is species A. Then redo your calculation, now using the birth data as well.

*First we need to determine the* 
$Pr(PandaSpecies_A|positiveTest) = \frac{Pr(positiveTest|PandaSpecies_A)*Pr(PandaSpecies_A)}{Pr(positiveTest)}$
$Pr(positiveTest) = Pr(positiveTest|PandaSpecies_A)*Pr(PandaSpecies_A) + Pr(positiveTest|PandaSpecies_A)*Pr(PandaSpecies_A)$
```{r}
paste("Pr(positiveTest) =",
      (0.8 * 0.5)+(0.35*0.5))


paste("Pr(PandaSpecies_A|positiveTest) =",
      (0.8 * 0.5)/(0.575))


```
First lets find the $Pr(singleton)$  
$Pr(singleton) = Pr(singleton|PandaSpecies_A)*Pr(PandaSpecies_A) + Pr(singleton|PandaSpecies_B)*Pr(PandaSpecies_A)$
```{r probabilityOfsingle}
paste("Pr(singleton) = appox",
      (0.9 * 0.5)+(0.8 * 0.5))
```



PandaA  
S,S = 0.9 * 0.9 = 0.81  
S,T = 0.9 * 0.1 = 0.09
T,S = 0.1 * 0.9 = 0.09  
T,T = 0.1 * 0.1 * 0.01

PandaB  
S,S = 0.8 * 0.8 = 0.64  
S,T = 0.8 * 0.2 = 0.16
T,S = 0.2 * 0.8 = 0.16  
T,T = 0.2 * 0.2 * 0.04

